---
title: "Adding italic species names to ggplot2 figures"
description: |
   How to add italics to ggplot figures
author:
  - name: Guy F. Sutton
    url: https://twitter.com/stats_ecology/
date: 09-17-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
draft: false
#base_url: 
categories:
  - R
  - tidyverse
  - ggplot2
  - ggtext
  - R markdown
  - PalmerPenguins
  - Data visualisation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tidyr,
               ggtext,
               palmerpenguins,
               here,
               cowplot) 
```

# Background 

The scared ecologist often finds themselves needing to make beautiful figures to include in their thesis or a new paper they want to submit to a high-profile journal. They open up `R`, load the `ggplot2` package, make their figure, and save it to their PC... Voila! However, they realise that they actually needed to italicise the names of the species they work on. For example, consider Figure 4 in [Towers and Dwyer, 2021](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1002/2688-8319.12043), whereby the relationship between plant basal area and stem diameter are plotted for two different plant species (*Acacia harpophylla* and *Casuarina cristata*). This often results in many hours of trawling the internet trying to figure out how to do this in `ggplot2`, often with little success. They then resort to spending hours trying to annotate their plots in programmes such as InkScape or Adobe Illustrator. 

In today's blogpost, I am going to demonstrate a simple way to add italicised species names to your beautiful `ggplot2` graphics, and make simple `ggplot2` plots look a little neater, without needing to export your figures into other software. 

Let's load the [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/) dataset into `R`. This dataset contains body morphology measurements for a range of penguin species from the Antarctic islands ([Gorman et al. 2014](https://doi.org/10.1371/journal.pone.0090081)). 

```{r load_data}
# Load raw penguins data
penguins <- palmerpenguins::penguins %>%
  # Remove rows containing NA data
  tidyr::drop_na(body_mass_g)
head(penguins)
```

# Figure 1

Let's assume we have analysed our dataset, and we now want to plot some figures. Let's go ahead and make a basic figure, and we will incrementally make changes to the figure to make it publication ready.

### Step 1: Plot basic figure

In the first figure, we will plot the flipper length (in mm) on the x-axis and body mass (in g) on the y-axis, and colour the data points according to penguin species. 

```{r plot_figure1}
penguins %>%
  ggplot(data = ., aes(x = flipper_length_mm,
                       y = body_mass_g,
                       colour = species)) +
  geom_point()
```

### Step 2: Set default ggplot theme

The first thing I recommend is changing the default `ggplot2` theme to make the figures look a little neater and prettier (at least to my tastes!). There are a range of built-in themes that you can use (e.g. `theme_classic()`, `theme_bw`, ect...), but I like the theme below that I have manually written. 

```{r set_ggplot_theme}
# Change and set default ggplot theme
theme_set(theme_classic() +
            theme(panel.border = element_rect(colour = "black", 
                                              fill = NA),
                  axis.text = element_text(colour = "black"),
                  axis.title.x = element_text(margin = unit(c(2, 0, 0, 0), 
                                                            "mm")),
                  axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), 
                                                            "mm"))))
```

Now, remake the figure from above. Notice the difference in appearance? Whenever we make a new plot, the theme we set above will automatically be applied. Nifty. 

```{r plot_figure1_themed}
penguins %>%
  ggplot(data = ., aes(x = flipper_length_mm,
                       y = body_mass_g,
                       colour = species)) +
  geom_point()
```

### Step 3: Changing axis labels and adding a title

```{r figure1_edit_axis}
penguins %>%
  ggplot(data = ., aes(x = flipper_length_mm,
                       y = body_mass_g,
                       colour = species)) +
  geom_point() +
  # Change the labels of axes and legend
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    colour = "Penguin species",
    title = "Do larger penguins (Pygoscelis spp.) weigh more?"
  )
```

### Step 4: Adding italics to title

While we have always been able to add italics to a `ggplot2` figure, the code has always been pretty messy and confusing, in my opinion. It would take me many attempts to finally get the italics in the correct place... Maybe it is just me, but I don't think so... 

Enter the amazing [ggtext](https://github.com/wilkelab/ggtext) package. This package allows us to use markdown language and/or HTML to edit our `ggplot2` figures by using the function `element_markdown()`. If you don't know any markdown, don't worry. There are many nice [introductions to markdown](https://rmarkdown.rstudio.com/lesson-1.html), but even better, to italise any text in our `ggplot2` figure, we simply need to put that text within a * either side of the word or phrase. For example, to make the word 'Panthera' italised, we need to type `*Panthera*`. 

Let's test it out and make the genus name for our penguins (*Pygoscelis*) italicised in our figure. 

```{r figure1_italics_title}
penguins %>%
  ggplot(data = ., aes(x = flipper_length_mm,
                       y = body_mass_g,
                       colour = species)) +
  geom_point() +
  # Change the labels of axes and legend
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    colour = "Penguin species",
    # We place * ___ * around the word we want italicised
    title = "Do larger penguins (*Pygoscelis* spp.) weigh more?"
  ) +
  # Use ggtext to specify that the axis title should be interpreted as markdown text
  theme(
    plot.title = ggtext::element_markdown()
  )
```

This looks like a pretty good figure to me. Let's set this aside and pat ourselves on the back for a figure well made. :) 

# Figure 2

Let's now make a second figure to demonstrate the power of `element_markdown()` and demonstrate how to add italicised text to our axis labels and figure legend. As above, we will go ahead and make a basic figure, and then incrementally make changes to the figure to make it publication ready.

### Step 1: Plot basic figure

In this second figure, we will plot the body mass (in g) on the y-axis and the penguin species on the x-axis, and colour the data points according to penguin species. We will use the `scale_fill_grey` function to colour the species-specific boxplots in greyscale. The start and end options specified within `scale_fill_grey` below make the shades lightgrey instead of dark grey/black (if you want darker shades, change this to: start = 0.2, end = 0.6). 

```{r}
# Make a basic graph
penguins %>%
  ggplot(data = ., aes(x = species,
                       y = body_mass_g,
                       fill = species)) +
  geom_boxplot() +
  # Make the boxplots different shades of grey
  scale_fill_grey(start = 0.5, end = 0.9) +
  labs(
    x = NULL,
    y = "Body mass (g)",
    fill = "Penguin species",
  ) 
```

### Step 2: Add species names to x-axis labels 

Now, let's add the genus and species names to the x-axis along with the common name of each penguin. For example, the species name of the Adelie penguin is *Pygoscelis adeliae* or *P. adeliae*, the Chinstrap penguin is *P. antarcticus*, and the Gentoo penguin is *P. papua*. There are many ways to do this. We are going to manually change the names in the raw data before plotting (you could also use `scale_x_discrete`, if you really wanted). 

```{r figure2_add_species}
# Now italicise and place species on own line (with fill)
penguins %>%
  # Here, we loop through the species names, and if the species name == x, change it to the name given to the right of ~
  dplyr::mutate(species = dplyr::case_when(
    # e.g. If species = 'Adelie', change to 'Adelie (P. adeliae)'
    species == "Adelie" ~ "Adelie (P. adeliae)",
    species == "Chinstrap" ~ "Chinstrap (P. antarcticus)",
    species == "Gentoo" ~ "Gentoo (P. papua)"
  )) %>%
  ggplot(data = ., aes(x = species,
                       y = body_mass_g,
                       fill = species)) +
  geom_boxplot() +
  scale_fill_grey(start = 0.5, end = 0.9) +
  labs(
    x = NULL,
    y = "Body mass (g)",
    fill = "Penguin species"
  )
```

### Step 3: Move species name to a new line below common name 

The x-axis labels look a bit silly in the above graph. It looks so messy having the common and species names on the same line. Let's move the species name to its own line below the common name in the x-axis label. Welcome back `element_markdown()`. 

We are going to leverage the power of using `element_markdown()` to use HTML to insert a linebreak between the common name and species name. To insert a linebreak in HTML, we just need to add `<br>` wherever we want the linebreak. 

```{r figure2_new_line}
# Use HTML linebreaks to split common and species names in x-axis labels
penguins %>%
  # Add <br> where you want the linebreak to occur
  dplyr::mutate(species = dplyr::case_when(
    species == "Adelie" ~ "Adelie <br> (P. adeliae)",
    species == "Chinstrap" ~ "Chinstrap <br> (P. antarcticus)",
    species == "Gentoo" ~ "Gentoo <br> (P. papua)"
  )) %>%
  ggplot(data = ., aes(x = species,
                       y = body_mass_g,
                       fill = species)) +
  geom_boxplot() +
  scale_fill_grey(start = 0.5, end = 0.9) +
  labs(
    x = NULL,
    y = "Body mass (g)",
    fill = "Penguin species"
  ) +
  # Use ggtext to specify that the x-axis text and legend text should be interpreted as markdown text
  theme(
    axis.text.x = element_markdown(),
    legend.text = element_markdown()
  )
```

### Step 4: Make species names into italics 

Much like for Figure 1, we can now make the species names italicised using `* ___ *` and `element_markdown()`. 

```{r figure2_italics, preview = TRUE}
# Make species names italicised 
penguins %>%
  # Remember, to make a word/phrase italic, put it in * _____ * 
  dplyr::mutate(species = dplyr::case_when(
    species == "Adelie" ~ "Adelie <br> (*P. adeliae*)",
    species == "Chinstrap" ~ "Chinstrap <br> (*P. antarcticus*)",
    species == "Gentoo" ~ "Gentoo <br> (*P. papua*)"
  )) %>%
  ggplot(data = ., aes(x = species,
                       y = body_mass_g,
                       fill = species)) +
  geom_boxplot() +
  scale_fill_grey(start = 0.5, end = 0.9) +
  labs(
    x = NULL,
    y = "Body mass (g)",
    fill = "Penguin species"
  ) +
  # Use ggtext to specify that the x-axis text and legend text should be interpreted as markdown text
  theme(
    axis.text.x = element_markdown(),
    legend.text = element_markdown()
  ) +
  # Increase the space between legend items (looks a bit nicer)
  theme(
    legend.key.height = unit(1.2, 'cm')
    )
```

```{r plot_save, echo = FALSE}
# Save qqplot above to disc
ggsave(here::here("./_posts/2021-09-17-adding-italics-to-ggplot-figures/penguin_plot_italics.png"),
       dpi = 600,
       height = 6,
       width = 8)
```

# Conclusions 

There we go. The amazing `ggtext` package makes it really easy to use simple markdown and HTML to quickly add italicised species names to our `ggplot2` figures and make them publication-ready quality without the need for editing figures in other software programmes. Are there other editing steps that you would performing outside of `R` to make your figures ready? Let me know and I am sure we can find a way to automate the process in `R`. 


