---
title: " "
output:
  pdf_document: default
  html_document: default
  word_document: default
---

\pagebreak

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\noindent

**Supplementary Materials: Enter name of your paper here**

 \  
 
Sutton, G. F.^1^$\dag$

 \  

\begingroup
\fontsize{10}{12}\selectfont
^1^ Center for Biological Control, Department of Zoology and Entomology, Rhodes University, Makhanda, 6140, South Africa
\endgroup

\

**Corresponding author** $\dag$ g.sutton@ru.ac.za

\pagebreak

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required packages
if (!require("pacman"))
  install.packages("pacman")
pacman::p_load(
  tidyverse,      # Basic data pre-processing and cleaning
  glmmTMB,        # Download and use the 'Salamanders' dataset
  DHARMa,         # Model residual diagnostics
  pander,         # Format R markdown documents 
  broom           # Extract tidy model summaries and coefficients
)

# Load the Salamanders dataset from glmmTMB package 
df <- glmmTMB::Salamanders %>%
  janitor::clean_names()
head(df)
```


```{r include = FALSE, echo = FALSE}
# Specify model formulae
formulae <- paste(
  # Specify the response variable here
  "count ~",
  # Specify the X predictor variables to fit the each model here
  # - Each line will be a different model
  c(# Model #1: count ~ wtemp
    "wtemp",
    # Model #2: count ~ doy
    "doy",
    # Model #3: count ~ wtemp + doy
    "wtemp + doy",
    # Model #4: count ~ wtemp * doy
    "wtemp * doy"))
formulae
```

```{r, warning = FALSE, message = FALSE, include = FALSE, echo = FALSE}
# Extract DHARMa
models <- tibble(formulae) %>%
  dplyr::mutate(
    # Run Poisson GLM's for each formula defined above in 'formulae' 
    mods = purrr::map(formulae,
                      ~ glm(
                        as.formula(.),
                        family = poisson(link = "log"),
                        data = df
                      )),
    # Extract DHARMa residuals plots for each model
    # - Note how we now pass 'purrr' the 'mods' column we created 
    #   containing a list of the models we have specified
    # - We are no longer using the formulae, but the resulting models 
    resid_plots = purrr::map(mods,
                             ~ DHARMa::simulateResiduals(., plot = F)),
    # Extract AICc for each model 
    AICc = purrr::map_dbl(mods,
                        ~ MuMIn::AICc(.))) %>%
    # Calculate delta AICc
  dplyr::ungroup() %>%
  dplyr::mutate(deltaAICc = round(AICc - min(AICc), digits = 2)) %>%
  # Process the model names to make the titles for each page cleaner
  dplyr::group_by(formulae) %>%
  dplyr::mutate(model_no = dplyr::cur_group_rows()) %>%
  # Here, we paste a bunch of information from different columns in our 
  # dataset and character strings (e.g. "...") to create a title 
  # for each page with the model number, the model formula and the 
  # delta AICc value of the model 
  dplyr::mutate(formulae = paste0("Model #", 
                                  model_no, 
                                  ": ", 
                                  formulae,
                                  " (Delta AICc = ",
                                  deltaAICc,
                                  ")"))

```

\newpage

# Appendix S1: Model diagnostics

```{r, results = "asis", warning = FALSE, message = FALSE, echo = FALSE, fig.width = 8}
# Automate the plotting of residual plots 
for (i in 1:nrow(models)) {
  
  # Add a title for each page
  pander::pandoc.header(models$formulae[i], level = 2)
  
  # Add the plot for each page
  plot(models$resid_plots[[i]]) 
  
  # Add a pagebreak between each model plot
  # - Note the extra '\' needed to escape the '\newpage' call function
  pander::pandoc.p('\\newpage')
  
}
```